Fail Faster

### 2025-03-15 paper/paper.tex
smithing words about alloc

The precise details of how the SplitMix algorithm works \cn are unimportant for the
present paper, but the critical component is that all of its operations are are
defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
This has a significant impact on the performance of generators. A single call to
one of the \bq library functions --- like generating an integer uniformly in a range --- may call the \texttt{splittable\_random}
algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
words \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
to the next garbage collection pause. While allocation in OCaml is fast and the garbage colletor
runs incrementally in small slices
\footnote{The OCaml GC is a generational collector \cn, and since these allocations are mostly very short lived, they will never be promoted. }, we will see in Section~\ref{section:eval} that this can have
a large performance impact on some generators that spend most of their time sampling data.
To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.

>>>>

The precise details of how the SplitMix algorithm works \cn are unimportant for the
present paper, but the critical component is that all of its operations are are
defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
This has a significant impact on the performance of generators. A single call to
one of the \bq library functions --- like generating an integer uniformly in a range --- may call the \texttt{splittable\_random}
algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
times \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently collect in OCaml,
\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}, we will see in Section~\ref{section:eval} that this can have
a large performance impact on some generators that spend most of their time sampling data.
To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.
